---
output: html_document
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, eval = FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```


# Background
Clear and appealing vizualization is essential to communicate scientific results. One of the main advantages of R is that it makes it easy to visualize data. In the last years, especially the grmmar of graphics framework has produced many different visualization methods.

# Objective
This exercise will provide you with scripts tpo plot speces richness maps and ancestral area reconstructions from BioGeoBEARS.

# Exercise


# Libraries

```{r}
library(ape)
library(ggtree)
library(tidyverse)
library(geiger)
library(stringr)
library(RColorBrewer)
library(colorspace)
library(jpeg)
library(viridis)
```

# Tutorial

## Plotting species richness maps

## PLotting results from INfomap bioregions

```{r}
#recode the bioregions names
bior.plot$id <- recode_factor(bior.plot$id,  
       "1 0" = "Northern South America",
       "2 0" = "Caribbean",
       "3 0" = "Central South America",
       "4 0" = "Africa",
       "5 0" = "Eastern South America",
       "6 0" = "Australasia",
       "7 0" = "North America",
       "8 0" = "Madagascar"
       )

bior.plot$id <- factor(bior.plot$id ,levels(bior.plot$id )[c(7, 2, 1, 3, 5, 4, 8, 6)])

proj4string(landmass) <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
landmass <- spTransform(landmass, w3)
mapWorld <- fortify(landmass)

cols <- brewer.pal(n = 8, name = "Set1")#[-9]

#to lump SA
#cols[c(3:5)] <- cols[3]


ggplot()+
  #mapWorld+
  geom_polygon(data = mapWorld, aes(x = long, y = lat, group = group), fill = "grey80")+
  geom_polygon(data = bior.plot, aes(x = long, y = lat, group = group, fill = id))+
  scale_fill_manual(values = cols)+
  #coord_map("cylequalarea", parameters = 30, xlim=c(-180, 180))+
  theme_bw()+
  coord_fixed()+
  #coord_map("cylequalarea", parameters = 30, xlim=c(-180, 180))
  theme(
    legend.position = c(0,0),
    legend.title = element_blank(),
    legend.justification = c(-0.3,-0.3),
    legend.text = element_text(size = 16),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )

```

# Plotting ancestral area reconstruction

## Load the data
```{r}
# The range probabilities
range_probabilities <-  read.table("example_data/bombacoideae_DEC_node_probabilities.txt", header = TRUE)

# The phylogenetic tree
tr <- read.tree("example_data/bombacoideae_phylogeny.tre")

# The tip classification used for the DEC model 
traits <-  read.table("example_data/bombacoideae_biome_classification.txt", header = TRUE, skip = 1)

# A map of the biomes
map <- readJPEG("example_data/Fig1b_bioregionalization_reduced.jpg")
```

## Set the paramaeters for the plotting
```{r}
n_states <- 5

state_names <- c("A", "B", "C", "D", "E")
```


```{r}
#node states
states <- range_probabilities%>%
  select(-c(1))%>%
  mutate(node = parse_number(rownames(.)))



#only keep the  most likely states and fuse others to "others"
plo.states <- list()
for(i in 1:nrow(states)){
  sub <- data.frame(t(slice(states, i)))
  ord <- rev(order(sub[,1]))
  out <- data.frame(t(sub[ord,][1:2]))
  names(out) <- rownames(sub)[ord][1:2]
  plo.states[[i]] <- round(out,3)
}

plo.states <- bind_rows(plo.states)
plo.states[is.na(plo.states)] <- 0
plo.states$OTHERS <- 1-rowSums(plo.states[,-1])
plo.states <- plo.states[-c(1:Ntip(tr)),-1]
plo.states <- plo.states[, colSums(plo.states) > 0]
areas <- names(plo.states)
names(plo.states) <- seq(1:ncol(plo.states))
# write.csv(plo.states, "output/node.states.csv")
# plo.states <- read_csv("output/node.states.csv")

#tip states
tip.states <- traits
names(tip.states) <- c("id", "a")
tips <- data.frame(row.names= tip.states$id,
                   str_split_fixed(string = tip.states$a, n = n_states, pattern = ""))
names(tips) <- state_names

brc <- c(viridis(n = n_states), rainbow(n = (ncol(states) - n_states)))

for(i in 1:n_states){
  tips[,i] <- as.numeric(tips[,i])
  tips[tips[,i] != 1, i] <- "grey90"
  tips[tips[,i] == 1, i] <- brc[i]
}
rownames(tips) <-tip.states$id

# #new tip labels
# tl <- read.csv("input/FINAL_Bombacoids_traits_input_1707.csv")[, c("id", "species")]
# rownames(tl) <- tl$id
# tl <- tl[tr$tip.label,]
# 
# tr$tip.label <- gsub("_", " ", as.character(tl$species))
# 
# tlp <- tl[tip.states$id,]
# rownames(tips) <- gsub("_", " ", as.character(tlp$species))


#pick nodes for pie label size
#from revells blog
getDescendants<-function(tree,node,curr=NULL){
  if(is.null(curr)) curr<-vector()
  daughters<-tree$edge[which(tree$edge[,1]==node),2]
  curr<-c(curr,daughters)
  w<-which(daughters>=length(tree$tip))
  if(length(w)>0) for(i in 1:length(w))
    curr<-getDescendants(tree,daughters[w[i]],curr)
  return(curr)
}

#geological epochs
epo.lim <- data.frame(x = c(-2.58, -5.33, -23.03, -33.09, -56, -66))

epo.lab <- data.frame(x = c(-2.58/2, -2.58 - (5.33-2.58) / 2 ,
                            -5.33 - (23.03 - 5.33) / 2 , 
                            -23.03 - (33.09 - 23.03) / 2 , 
                            -33.09 - (56 - 33.09) / 2,
                            -56 - (66 - 56) / 2,
                            -66 - (max(node.depth.edgelength(tr) - 66) / 2)),
                      lab = c("Q.", "Pl.", "Miocene", "Oligocene", "Eocene", "Paleocene", "Late Cretaceous"))

epo.lab <- epo.lab[1:5,]

epo.lim <- epo.lim + max(node.depth.edgelength(tr))
epo.lab$x <- epo.lab$x + max(node.depth.edgelength(tr))

#check the nodes of which the descendants should be small instead of large
plot(tr, cex = 0.5)
nodelabels()

sno <- c(getDescendants(tr, 189), getDescendants(tr, 167), getDescendants(tr,127), getDescendants(tr, 114),
         getDescendants(tr, 118), 107, 111, 112, 163, 164)
sno <- sno[!sno %in% 1:Ntip(tr)]


lno <- as.numeric(rownames(plo.states)[!as.numeric(rownames(plo.states)) %in% sno])
lno <- lno[!lno %in% 1:Ntip(tr)]
#lno <- lno[!lno %in% c(114,116, 117, 115, getDescendants(tr, 117))]#exclude the root and outgroup

#mix node colors
names(brc)[c(1:n_states)] <- state_names
node.col <- brc[areas]
names(node.col) <- areas

rains <- brc[-c(1:n_states)]
node.col[is.na(node.col)] <- rains
node.col["OTHERS"] <- "grey50"

#colours for the legend
leg <- node.col[order(nchar(names(node.col)))]

names(leg) <- c(names(states)[-length(names(states))], "OTHERS")

#base plotting
plot(tr, label.offset = 9.5, x.lim = 116, cex = 0.6, y.lim = 107)
segments(x0 = epo.lim$x, y0 = -8, x1 = epo.lim$x, y1 = 105, col = "grey50")
text(x = epo.lab$x, y = -1.5, labels = epo.lab$lab, cex = 0.6)
#tip labels
tiplabels(pch = 22, bg = tips[tr$tip.label,1], adj = 3.5, col = tips[tr$tip.label,1], cex = 0.8)
tiplabels(pch = 22, bg = tips[tr$tip.label,1], adj = 4.5, col = tips[tr$tip.label,1], cex = 0.8)
tiplabels(pch = 22, bg = tips[tr$tip.label,2], adj = 5.5, col = tips[tr$tip.label,2], cex = 0.8)
tiplabels(pch = 22, bg = tips[tr$tip.label,2], adj = 6.5, col = tips[tr$tip.label,2], cex = 0.8)
tiplabels(pch = 22, bg = tips[tr$tip.label,3], adj = 7.5, col = tips[tr$tip.label,3], cex = 0.8)
tiplabels(pch = 22, bg = tips[tr$tip.label,3], adj = 8.5, col = tips[tr$tip.label,3], cex = 0.8)
tiplabels(pch = 22, bg = tips[tr$tip.label,4], adj = 9.5, col = tips[tr$tip.label,4], cex = 0.8)
tiplabels(pch = 22, bg = tips[tr$tip.label,4], adj = 10.5, col = tips[tr$tip.label,4], cex = 0.8)

# tip label heading
text(labels = c("MBF", "DBF", "SAV"),
     x = seq(60, 64, by = 2), y = 104, cex = 0.5, srt = 90, adj = 0)
par(fg = "transparent")

# node label pie-charts
nodelabels(pie= as.matrix(plo.states[as.numeric(rownames(plo.states)) %in% lno,]),
           node = lno, piecol=node.col, cex = 1)
#for whatever reason this works, but vectorized nodelabels does not
for(i in 1:length(sno)){
  print(i)
  nodelabels(pie= as.matrix(plo.states[as.numeric(rownames(plo.states)) %in% sno[i],]),
             node = sno[i], piecol=node.col, cex = 0.5)
}

par(fg = "black")
axisPhylo(1)
rasterImage(map,
            xleft = -5,
            ybottom = 95,
            xright = 45 ,
            ytop = 90 + 50 * (615/1027) * (8 / 11))

legend(x = -5, y = 100, ncol = 1, legend = names(leg),
       fill = leg, cex = 0.9, bty = "n")

```