---
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, eval = FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```

In this exercise you will run the DIspersal-Extinction-Cladogenesis model using the BioGeoBEARS R package.

# Library setup

```{r}
library(tidyverse)
library(ape)
library(geiger)
library(optimx)         # You need to have some version of optimx available
library(FD)       # for FD::maxent() (make sure this is up-to-date)
library(snow)     # (if you want to use multicore functionality; some systems/R versions prefer library(parallel), try either)
library(parallel)
library(devtools)
library(rexpokit)
library(cladoRcpp)
library(BioGeoBEARS)
library(stringr)
library(RColorBrewer)
library(colorspace)
library(jpeg)
library(viridis)
```

# Tutorial
This script is mostly absed on the example script by the developer of BioGeoBEARS, available at http://phylo.wikidot.com/biogeobears.

## Prepare the tree file
If your tree is in nexus format, you need to convert it to newick format, otherwise you can jus specify the file apth
```{r}
#specify the tree file
trfn <- "example_data/bombacoideae_phylogeny.newick"
tr <- read.tree(trfn)
```

## Define the grography file
```{r}
geogfn <- "example_data/bombacoideae_biome_classification.txt"

# Look at your geographic range data:
tipranges = getranges_from_LagrangePHYLIP(lgdata_fn=geogfn)
```

## Set up the DEC model

```{r}
BioGeoBEARS_run_object = define_BioGeoBEARS_run()

# set the macimum number of areas. If you ahve many areas, you can set the amximum number of areas a species can occur in simultaneously. One possibility is to set this to the number of areas of the most widespread recent species.
max_range_size = 5 
numstates_from_numareas(numareas=5, maxareas=5, include_null_range=TRUE)

# Set some operators for the default DEC model

BioGeoBEARS_run_object$trfn = trfn
BioGeoBEARS_run_object$geogfn = geogfn
BioGeoBEARS_run_object$max_range_size = max_range_size
BioGeoBEARS_run_object$min_branchlength = 0.000001    
BioGeoBEARS_run_object$include_null_range = TRUE  
BioGeoBEARS_run_object$speedup = TRUE          #
BioGeoBEARS_run_object$use_optimx = 'GenSA'     # if FALSE, use optim() instead of optimx()
BioGeoBEARS_run_object$num_cores_to_use = 4
BioGeoBEARS_run_object$force_sparse = FALSE    
BioGeoBEARS_run_object = readfiles_BioGeoBEARS_run(BioGeoBEARS_run_object)
BioGeoBEARS_run_object$return_condlikes_table = TRUE
BioGeoBEARS_run_object$calc_TTL_loglike_from_condlikes_table = TRUE
BioGeoBEARS_run_object$calc_ancprobs = TRUE    # get ancestral states from optim run
runslow = TRUE
```

## Define the output path
```{r}
resfn = "example_data/bombacoideae_DEC_results.Rdata"
```

## Check if the the input is alright
```{r}
check_BioGeoBEARS_run(BioGeoBEARS_run_object)
```

## Run the DEC model
```{r}
resDEC = bears_optim_run(BioGeoBEARS_run_object)
```

## Save outputs to disk
```{r}
# Save the result file
save(resDEC, file=resfn)

# Save the node states for visualization
resDEC$ML_marginal_prob_each_state_at_branch_top_AT_node
trtable = prt(tr, printflag=FALSE)
areas = getareas_from_tipranges_object(tipranges)
states_list_0based = rcpp_areas_list_to_states_list(areas=areas, maxareas=max_range_size, include_null_range=TRUE)

# Make the list of ranges
ranges_list = NULL
for (i in 1:length(states_list_0based))
{    
  if ( (length(states_list_0based[[i]]) == 1) && (is.na(states_list_0based[[i]])) )
  {
    tmprange = "_"
  } else {
    tmprange = paste(areas[states_list_0based[[i]]+1], collapse="")
  }
  ranges_list = c(ranges_list, tmprange)
}

range_probabilities = as.data.frame(resDEC$ML_marginal_prob_each_state_at_branch_top_AT_node)
row.names(range_probabilities) = trtable$node
names(range_probabilities) = ranges_list

# Write the table to a tab-delimited text file (for Excel etc.)
write.table(range_probabilities, file="example_data/bombacoideae_DEC_node_probabilities.txt",
            row.names = F, quote=FALSE, sep="\t")
```

## Visualization
```{r}

analysis_titletxt ="BioGeoBEARS DEC"

# Setup
results_object = resDEC
scriptdir = np(system.file("extdata/a_scripts", package="BioGeoBEARS"))

# A plot of ancestral states
res2 = plot_BioGeoBEARS_results(results_object, analysis_titletxt, 
                                addl_params=list("j"), plotwhat="text", 
                                label.offset=2, tipcex=0.7, statecex=0.6, 
                                splitcex=0.6, titlecex=0.8, plotsplits=F, 
                                cornercoords_loc=scriptdir, include_null_range=TRUE, 
                                tr=tr, tipranges=tipranges, plotlegend = T)

# showing pie charts for each node with the probabilities
plot_BioGeoBEARS_results(results_object, analysis_titletxt, 
                         addl_params=list("j"), plotwhat="pie", 
                         label.offset=2, tipcex=0.7, statecex=0.7, 
                         splitcex=0.6, titlecex=0.8, plotsplits=F, 
                         cornercoords_loc=scriptdir, include_null_range=TRUE, 
                         tr=tr, tipranges=tipranges)
```

Done!