---
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, eval = FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```

In this exercise, you will use the results of the ancestral state recosntruction, to do a biogeographic stochastic mapping, and count the number of shifts throught time.

# Library setup
```{r}
library(tidyverse)
library(ape)
library(geiger)
library(optimx)         # You need to have some version of optimx available
library(FD)       # for FD::maxent() (make sure this is up-to-date)
library(snow)     # (if you want to use multicore functionality; some systems/R versions prefer library(parallel), try either)
library(parallel)
library(devtools)
library(rexpokit)
library(cladoRcpp)
library(BioGeoBEARS)
library(stringr)
library(RColorBrewer)
library(colorspace)
library(jpeg)
library(viridis)
```



# Loading the results of the DEC ancestral area reconstruction

```{r}
load("example_data/bombacoideae_DEC_results.Rdata")
model_name = "DEC"
results_object = res = resDEC
scriptdir = np(system.file("extdata/a_scripts", package="BioGeoBEARS"))

clado_events_tables = NULL
ana_events_tables = NULL
lnum = 0

BSM_inputs_fn = "BSM_inputs_file.Rdata"
runInputsSlow = TRUE
if (runInputsSlow)
{
  stochastic_mapping_inputs_list = get_inputs_for_stochastic_mapping(res=res)
  save(stochastic_mapping_inputs_list, file=BSM_inputs_fn)
} else {
  # Loads to "stochastic_mapping_inputs_list"
  load(BSM_inputs_fn)
} # END if (runInputsSlow)
```

# Run the biogeographic stochastic mapping

```{r}
set.seed(seed=as.numeric(Sys.time()))

runBSMslow = TRUE

BSM_output = runBSM(res,
                    stochastic_mapping_inputs_list = stochastic_mapping_inputs_list, 
                    maxnum_maps_to_try=2000, 
                    nummaps_goal=10, # INcrease the number of replicates if possible
                    maxtries_per_branch=40000,
                    save_after_every_try=TRUE, 
                    savedir=getwd(), 
                    seedval=12345,
                    wait_before_save=0.01)

RES_clado_events_tables = BSM_output$RES_clado_events_tables
RES_ana_events_tables = BSM_output$RES_ana_events_tables
```

# Save the results to disk
```{r}
# Extract BSM output
clado_events_tables = BSM_output$RES_clado_events_tables
ana_events_tables = BSM_output$RES_ana_events_tables
save(clado_events_tables, file = "example_data/bombacoideae_BSM_clado_events_table_mcc_detbiom.rda")
save(ana_events_tables, file ="example_data/bombacoideae_BSMana_events_table_mcc_detbiom.rda")
```

## Get the number of shifts through time
You may want to adapt the maximum age here!
```{r}
angen <- lapply(ana_events_tables, function(k){
  out <- k %>% 
  dplyr::select(abs_event_time, event_type, event_txt, dispersal_to) %>% 
  mutate(bin = cut(abs_event_time, 
                   breaks = seq(24, 0, by = -2), 
                   labels = rev(seq(23, 1, by = -2))))%>% 
    mutate(from = str_split_fixed(event_txt, n = 2, pattern = "->")[,1 ]) %>% 
    mutate(sdb_efb = ifelse(!grepl("W", from) & grepl("W", dispersal_to), 1, 0)) %>% 
    mutate(efb_sdb = ifelse(from == "W" & (grepl("D", dispersal_to) | grepl("S", dispersal_to)), 1, 0))
  return(out)
  })

# cladogenetic events do not include dispersal to new biomes
cladgen <- lapply(clado_events_tables, function(k){
  out <-  k %>% 
  dplyr::select(time_bp, clado_event_type, clado_event_txt, clado_dispersal_to) %>% 
  filter(!is.na(clado_event_type) & clado_event_type != "")
  return(out)
})
```

# Count the number of shifts per time bin
Again you may need to adapt the maximum age and the number of areas.
```{r}
disps <- lapply(angen, function(k){
  out <-  k %>% 
    filter(event_type == "d") %>%
    group_by(bin, dispersal_to) %>% 
    summarize(abs_count = n()) %>% 
    ungroup() %>% 
    mutate(bin = parse_number(as.character(bin)))
  
  fram <- expand_grid( bin = rev(seq(23, 1, by = -2)), 
                       dispersal_to = c("A", "B", "C", "D", "E"))
  out <- left_join(fram, out, by = c("dispersal_to", "bin")) %>% 
    replace_na(list(abs_count = 0))
})

disps <- bind_rows(disps,.id = "SM") %>% 
  group_by(bin, dispersal_to) %>% 
  summarize(lower = quantile(abs_count, probs = 0.025),
            mean = mean(abs_count),
            upper = quantile(abs_count, probs = 0.975))
```

# Get the total branch length per time bin
```{r}
agegap = 2
branchtimes <-  branching.times(tr)
maxmilcatogories <-  ceiling(max(branchtimes) / agegap)
milgaps_mod <- matrix(ncol = 2, nrow = maxmilcatogories, 0)
colnames(milgaps_mod) <- c("time","tot_brl")
milgaps_mod[,1] <- 1:maxmilcatogories*agegap

## Sum times per bin
v <- branchtimes
for(i in 1:length(milgaps_mod[, 2])){
  s <- milgaps_mod[i, 1]        # time start
  e <- milgaps_mod[i, 1] - agegap # time end
  l1 <- length(v[v > s]) * agegap + min(length(v[v > s]), 1) * agegap
  l2 <- sum(v[v > e & v<s] -e)
  
  if (l1 == 0){
    l2 = l2*2
  } # root
  
  milgaps_mod[i,2] <- l1 + l2
}

brls <- data.frame(milgaps_mod) %>% 
  mutate(time = time - 1)
```

# Normalize the number of shifts by the total amount of branch length per time bin
```{r}
disps <- left_join(disps, brls, by = c("bin" = "time")) %>% 
  mutate(lower = lower / tot_brl) %>% 
  mutate(mean = mean / tot_brl) %>% 
  mutate(upper = upper / tot_brl)
```

```{r}
ggplot()+
  geom_point(data = disps, aes(x = bin, y = mean, color = dispersal_to))+
  geom_line(data = disps, aes(x = bin, y = mean, color = dispersal_to))+
  theme_bw()
```


