---
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, eval = FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```


# Joint Bayesian estimation of speciation, dispersal and extinction rates: the GeoSSE model

The GeoSSE model jointly infers area-specific speciation rates, dispersal rates and local extinction rates [(Goldberg et al. 2011)](https://academic.oup.com/sysbio/article/60/4/451/1611105). The `mcmc-SSE` program implements a Bayesian algorithm to infer these rates. 

The analysis requires a tree file, a table with the geographic ranges (species with no geographic data or not present in the tree will be automatically removed).


# Load dependent libraries: 
```{r}
library("diversitree")  
library("picante")
```

# Load mcmc-SSE library
You will find the mcmc-SSE-lib file in the "example_file" dropbox folder.
```{r}
source("my_path/mcmc-SSE-lib.R")
```

# Prep data
Since the GoeSSE model only allows for 2 areas the input data might need to be modified by merging some of the areas.

Define which areas should be merged

```{r}
data_SPGC <- read.csv("example_data/bombacoideae_biome_classification_details.csv", row.names = 1)
columns_areaA = c(1, 3)
columns_areaB = c(2, 4, 5)

geosse_areas <- comb_areas(data_SPGC, columns_areaA, columns_areaB)
write.table(geosse_areas, file="example_data/traitGeoSSE_owndataconverted.txt", sep="\t", row.names=F, quote=F)
```

# Setup an analysis
Set working directory:  
```{r}
setwd("my_path/example_files/")
```

Define tree file (NEXUS format, can contain multiple trees)  
```{r}
tree_file  = "example_data/bromelioieae_consensus.tre"
```



Define geographic range file (tab-separated text file). Geographic ranges are defined as 0 (present in area A), 1 (present in area B), 2 (present in area AB)  
```{r}
range_file = "example_data/traitGeoSSE.txt"
```

State-specific taxon sampling (A, B, AB)  
```{r}
sampling_fractions = c(0.7,0.8,1)
```

Define output file name
```{r}
mcmc.log = "bromeliad_geosse.log"
```

Run the analysis
```{r}
run_mcmc_SSE(tree_file, 
             range_file, 
             model = "geosse",
             outfile = mcmc.log,
             iterations = 10000, 
             rho = sampling_fractions)
```

# Process and plot output
Read output file
```{r}
post = read.table(mcmc.log, header=T)
```

Plot trace of sampled likelihoods
```{r}
plot(post$likelihood, type="l")
```

Plot speciation, dispersal, and extinction rates per area:  

```{r}
par(mfrow=c(1,3))
boxplot(post[,c("sA","sB","sAB")], col = "blue", main = "speciation rates")
boxplot(post[,c("dA","dB" )], col = "gray", main = "dispersal rates")
boxplot(post[,c("xA","xB" )], col = "red",main = "extinction rates")
```

Are dispersals rates significantly different?  

```{r}
delta_dispersal = post$dA - post$dB
hist(delta_dispersal)
```

Calculate posterior probability of dAB > dBA:
```{r}
length(delta_dispersal[delta_dispersal>0]) / length(delta_dispersal)
```

Are extinction rates significantly different?

```{r}
delta_extinction = post$xA - post$xB
hist(delta_extinction)
```

Calculate the posterior probability of xA > xB:

```{r}
length(delta_extinction[delta_extinction>0]) / length(delta_extinction)
```


# Run on multiple trees
If provided with a tree file that includes multiple trees the MCMC can run sequentially on different trees to account for phylogenetic uncertainty. The `burnin` defines how many initial MCMC iterations should be discarded (i.e. not saved in the log file) for each tree.

```{r}
tree_file  = "bromelioieae_100.trees"  
run_mcmc_SSE(tree_file, trait_file, model= "geosse", outfile = "bromeliad_geosse100.log", iterations = 10000, nTREES=5, burnin=100)
```

